Subset coercions in \Coq

### 2006-07-18 papers/article/main.tex
CP fixes

%\usepackage{qzapfcha}
%\usepackage{cmbright}
\usepackage{concmath}
\usepackage{ifthen}
\usepackage{varioref}

@@

@@

\abstract{We set out to establish a phase distinction between writing and
  proving algorithms in the \Coq{} environment. Concretely, this means
  allowing to write algorithms as easily as in a pratical functionnal 
  programming language
  whilst giving them as rich a specification as desired and proving that
  the code meets the specification using the whole \Coq{} proof
  apparatus. This is achieved by extending conversion to an equivalence 
  which relates types
  and subsets based on them, a technique  
  originating from the ``\emph{Predicate subtyping}'' feature of \PVS{} and
  following mathematical convention. The typing judgments can be
  translated to the \CICfull (\CIC{}) by means of an interpretation which inserts
  coercions at the appropriate places. These coercions can contain
  existential variables representing the propositional parts of the final
  term. A prototype implementation of this process
  has been implemented.}

\section{Introduction}
There are many means to program in the \Coq{} environment. One
can write programs as in \ML{} and prove properties about them
separately, loosing the possibility of using dependent types in
specifications, or give a rich type expressing them as a goal and use the
proof tactics to solve it, producing a corresponding program by the
\CurryHoward isomorphism but having much less control on its algorithmic
essence. It is however difficult to mix the two methods
(writing code and prove interactively) using a rich specification. We
propose a solution to overcome this difficulty, based on the \ps{}
\cite{Shankar&Owre:WADT99} technique of \PVS{}
\cite{PVS:Language,PVS-Semantics:TR}. This solution also provides
facilities to express properties with a syntaxicaly more mathematical flavor,
bridging the gap between mathematical convention and type theory.

The \Program{} tactic by C. Parent \cite{conf/mpc/Parent95} had the same
goal as ours but a slightly different method for achieving it. It was strongly
linked to the extraction mechanism included in \Coq, both theoreticaly
and practically. Sketching the mechanism, they defined a weakened
extraction operation on \CIC{} terms which could be inverted because it left enough
information in the extracted term to rebuild a partial proof. The
mechanism, while general and theoreticaly well thought out, required
some heuristics and did not integrate smoothly with the \Coq{}
environment. In particular it lacked the pervasivness our method has,
being applicable in a wide variety of situations in the proof assistant
environment.

@@

The rest of the article is organized as follows: first we present a
type system based on \ECC{}, which integrates subset ``subtyping'', and
we show how it relates to \CIC{} by means of traductions between
judgements of the two systems. Then, we present a prototype
implementation with some case studies, and finally we give our
conclusions on this work and discuss future directions.

@@

    \AXC{$`G \type a : A$}
    \AXC{$`G \type p : P[a/x]$}
    \BIC{$`G \type \elt{A}{(\lambda x : A.P)}{a}{p} : \mysubset{x}{A}{P}$}
    \DP
    \vspace{\infvspace}

    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \UIC{$`G \type \text{proj1\_sig}~t : A$}
    \DP
    \vspace{\infvspace}

    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \UIC{$`G \type \text{proj2\_sig}~t : P[\text{proj1\_sig}~t/x]$}
    \DP
  \end{center}
  \vspace{-1em}
  \caption{Subset type in \CCI}
  \label{fig:cci-subset}
  \vspace{-2.5em}
\end{figure}

@@

\subsubsection{Equivalence}
We have renamed the technique from \ps{} to 
\emph{subset equivalence} because we have a symmetric relation, contrary
to usual subtyping relations. It also conveys the idea that it can
include the usual $\beta\pi$-conversion directly in the judgement, 
although it may not be possible if we have cummulative universes.
The judgment $`G \typed T \subd{} U : s$ (figure
\vref{fig:conversion-decl-rules}) reads $T$ is equivalent to $U$ in
environment $`G$, both being sorted by $s$.

@@

The rule \irule{SubConv} integrates $\beta\pi$-conversion in the
judgment. We use a judgemental equality here, which will be refined by
the usual conversion relation later. The \irule{SubSym} and
\irule{SubTrans} rules ensure that our judgement constructs an \emph{equivalence} and has
proper modularity. The next two rules do context closure for dependent
products and sums. It is remarkable that we use contravariance
\cite{journals/toplas/Castagna95} for
domains in the \irule{SubProd} rule, not restricting to invariance as in
\PVS. It is accessory here, as we could have used \emph{co}variance and
still get the same judgements because we have symetry. However, it will
become important when we create coercions (see
\vref{fig:coerce-impl-rules-short}).

@@

\begin{paragraph}{}
We use a standard transformation to get an algorithm from our typing
rules, integrating the conversion (or subsumption) rule
(here \irule{Coerce}) to the premises of the other rules.
We recall that $`G \type t : s$ is shorthand for $`G \type t : T$ and
$T  `= s$ with $s `: \PTSsorts$ in a algorithmic presentation.
\TODO{App?, SumDep?, Pi?}. Our algorithmic judgement $`G \typea t : T$
(figure \vref{fig:typing-algo-rules})
reads: $t$ is a well-typed term of type $T$ in environment $`G$.
We need to introduce the notion of support for subset types in order to
define our algorithmic system. Indeed, when typing an application, 
we need to ensure that the object we apply can be seen as an object of a
product type modulo the equivalence. This notion is presented in \PVS{} by
means of a $\mu_0$ operator which was readily adaptable in our setting
\vref{fig:mualgo-definition}. In Gang Chen's work, the notion is formalized by a
new judgement $\Pi_{lub}$ which is essentialy equivalent to the operator.
\begin{figure*}[t]
  \def\infvspace{0.5em}
    \def\type{\typea}
    \def\subt{\subta}
    \def\sub{\suba}
    \def\fCenter{\typea}
    \begin{center}
%     \WfEmpty\DP

@@

\end{paragraph}
\begin{paragraph}{}
  The next thing to do is to construct a decidable judgement for
  equivalence given two types. We denote it by $`G \typea T
  \suba U : s$ (figure \vref{fig:subtyping-algo-rules}).
  In our case, we have to check that conversion is decidable, which is the
  case. We also need to eliminate the transitivity rule which is not
  syntax-directed. That is the purpose of the following theorem:
\begin{theorem}[Elimination of transitivity]
  If $`G \typea T \suba U : s$ and $`G \typea U \suba V : s$, then $`G \typea T
  \suba V : s$.
\end{theorem}
\begin{proof}
  By induction on the sum of the depths of the two derivations.
  \TODO{example?}
\end{proof}
\end{paragraph}

@@

We can now prove the following:
\begin{theorem}[Forgetful map correctness]
  If $`G \typec t : T$ then $`G^{\circ} \typea t^{\circ} : T^{\circ}$ if $()^{\circ}$ is
  defined on $`G, t$ and $T$.
\end{theorem}

Practically, this means we can use almost any existing definition in the
\Coq environment, potentially turning them (locally) to an opaque form
if their body is not traductible. Obviously, this map is involutive and
the identity on \Russell terms.

\section{From \lng{} to \CCq{}}
We will now build an interpretation
$\ip{t}{`G}$ of \Russell terms $t$ in environment $`G$ into \CCI
terms. We will check at the end that it respects the $()^{\circ}$ operator
in the sense that if $t$ is well-typed in $`G$ then $\ip{t}{`G}^{\circ} = t$.

@@

The proof of this proposition is the main technical contribution of this
paper. It also represented the main difficulty in our work. 
The remainder of this section is organized as follows: first we will
define the explicit coercion derivation algorithm which will permit to
put proof obligations in the terms, then we will define the
interpretation of terms which is mutually recursive with the precedent
algorithm. Finaly we will present a proof sketch of the aforementionned
proposition.

@@

\subtqis[Coercion derivation]

>>>>

%\usepackage{qzapfcha}
%\usepackage{cmbright}
\usepackage{concmath}
\usepackage{ifthen}
\usepackage{url}
\usepackage{varioref}

@@

\def\CCI{\CIC}

@@

\abstract{We propose a new language for writing programs with dependent
  types which can be elaborated into \Coq terms. This language permits
  to establish a phase distinction between writing and proving
  algorithms in the \Coq{} environment.  Concretely, this means allowing
  to write algorithms as easily as in a pratical functionnal programming
  language whilst giving them as rich a specification as desired and
  proving that the code meets the specification using the whole \Coq{}
  proof apparatus. This is achieved by extending conversion to an
  equivalence which relates types and subsets based on them, a technique
  originating from the ``\emph{Predicate subtyping}'' feature of \PVS{}
  and following mathematical convention. The typing judgments can be
  translated to the \CICfull (\CIC{}) by means of an interpretation
  which inserts coercions at the appropriate places. These coercions can
  contain existential variables representing the propositional parts of
  the final term. A prototype implementation of this process has been
  implemented.}

\section{Introduction}
There are many means to program in the \Coq{} environment. One
can write programs as in \ML{} and prove properties about them
separately, loosing the possibility of using dependent types in
specifications, or give a rich type expressing them as a goal and use the
proof tactics to solve it, producing a corresponding program by the
\CurryHoward isomorphism but having much less control on its algorithmic
essence. It is however difficult to mix the two methods
(writing code and prove interactively) using a rich specification. We
propose a solution to overcome this difficulty, based on the \ps{}
\cite{Shankar&Owre:WADT99} technique of \PVS{}
\cite{PVS:Language,PVS-Semantics:TR}. This solution also provides
facilities to express properties with a more mathematical flavor,
bridging a gap between mathematical convention and type theory.

The \Program{} tactic by C. Parent \cite{conf/mpc/Parent95} had the same
goal as ours but a slightly different method for achieving it. It was strongly
linked to the extraction mechanism included in \Coq, both theoreticaly
and practically. Sketching the mechanism, she defined a weakened
extraction operation on \CIC{} terms which could be inverted because it left enough
information in the extracted term to rebuild a partial proof. The
mechanism, while general and theoreticaly well thought out, required
some heuristics and did not integrate smoothly with the \Coq{}
environment. In particular it lacked the pervasivness our method has,
being applicable in a wide variety of situations in the proof assistant
environment.

@@

The rest of the article is organized as follows: first we present a
type system based on \CC, which integrates subset ``subtyping'', and
we show how it relates to \CIC{} by means of traductions between
judgements of the two systems. Then, we present a prototype
implementation with some case studies, and finally we give our
conclusions on this work and discuss future directions.

@@

    \AXC{$`G \type a : A$}
    \AXC{$`G \type p : P[a/x]$}
    \AXC{$`G \type \mysubset{x}{A}{P} : \Set$}
    \TIC{$`G \type \elt{A}{(\lambda x : A.P)}{a}{p} : \mysubset{x}{A}{P}$}
    \DP
    \vspace{\infvspace}

    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \UIC{$`G \type \eltpit~t : A$}
    \DP
    \vspace{\infvspace}

    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \UIC{$`G \type \eltpip~t : P[\eltpit~t/x]$}
    \DP
  \end{center}
  \vspace{-1em}
  \caption{Subset type in \CCI}
  \label{fig:cci-subset}
  \vspace{-2.5em}
\end{figure}

@@

\subsubsection{Equivalence}
We have renamed the technique from \ps{} to 
\emph{subset equivalence} because we have a symmetric relation, contrary
to usual subtyping relations. It also conveys the idea that it can
include the usual $\beta\pi$-conversion directly in the judgement.
The judgment $`G \typed T \subd{} U : s$ (figure
\vref{fig:conversion-decl-rules}) reads $T$ is equivalent to $U$ in
environment $`G$, both being sorted by $s$.

@@

The rule \irule{SubConv} integrates $\beta\pi$-conversion in the
judgment. We use a judgemental equality here, which will be refined by
the usual conversion relation later. The \irule{SubSym} and
\irule{SubTrans} rules ensure that our judgement constructs an \emph{equivalence} and has
proper modularity. The next two rules (\irule{SubProd} and
\irule{SubSigma}) do context closure for dependent
products and sums. It is remarkable that we use contravariance
\cite{journals/toplas/Castagna95} for
domains in the \irule{SubProd} rule, not restricting to invariance as in
\PVS. It is accessory here, as we could have used \emph{co}variance and
still get the same judgements because we have symetry. However, it will
become important when we create coercions (see
\vref{fig:coerce-impl-rules-short}).

@@

\begin{paragraph}{}
We use a standard transformation to get an algorithm from our typing
rules, integrating the conversion (or subsumption) rule
(here \irule{Coerce}) to the premises of the other rules.
%We recall that $`G \type t : s$ is shorthand for $`G \type t : T$ and
%$T  `= s$ with $s `: \PTSsorts$ in a algorithmic presentation.
\TODO{App?, SumDep?, Pi?}. Our algorithmic judgement $`G \typea t : T$
(figure \vref{fig:typing-algo-rules})
reads: $t$ is a well-typed term of type $T$ in environment $`G$.
We need to introduce the notion of support for subset types in order to
define our algorithmic system. Indeed, when typing an application, 
we need to ensure that the object we apply can be seen as an object of a
product type modulo the equivalence. This notion is presented in \PVS{} by
means of a $\mu_0$ operator which was readily adaptable in our setting
(figure \vref{fig:mualgo-definition}). In Gang Chen's work, the notion is formalized by a
new judgement $\Pi_{lub}$ which is essentialy equivalent to the operator
$\mualgo$, except ours finds $\Sigma_{lub}$s too.
\begin{figure*}[t]
  \def\infvspace{0.5em}
    \def\type{\typea}
    \def\subt{\subta}
    \def\sub{\suba}
    \def\fCenter{\typea}
    \begin{center}
%     \WfEmpty\DP

@@

\end{paragraph}
\begin{paragraph}{}
  The next thing to do is to construct a decidable judgement for
  equivalence given two types. We denote it by $`G \typea T
  \suba U : s$ (figure \vref{fig:subtyping-algo-rules}).
  In our case, we have to check that conversion is decidable, which is the
  case. We also need to eliminate the transitivity rule which is not
  syntax-directed. That is the purpose of the following theorem:
\begin{theorem}[Admissibility of transitivity]
  If $`G \typea T \suba U : s$ and $`G \typea U \suba V : s$, then $`G \typea T
  \suba V : s$.
\end{theorem}
\begin{proof}
  By induction on the sum of the depths of the two derivations.
  \TODO{example?}
\end{proof}
\end{paragraph}

@@

We can now prove the following:
\begin{theorem}[Forgetful map correctness]
  If $`G \typec t : T$ then $`G^{\circ} \typea t^{\circ} : T^{\circ}$ if $()^{\circ}$ is
  defined (ie, does not return $`_$) on $`G, t$ and $T$.
\end{theorem}

Practically, this means we can use almost all existing definitions in the
\Coq environment. Obviously, this map is involutive and
the identity on \Russell terms. 
%, potentially turning them (locally) to an opaque form
% if their body is not traductible
\section{From \lng{} to \CCq{}}
We will now build an interpretation
$\ip{t}{`G}$ of \Russell terms $t$ in environment $`G$ into \CCI
terms. We will check at the end that it respects the $()^{\circ}$ operator
in the sense that if $t$ is well-typed in $`G$ then $\ip{t}{`G}^{\circ} = t$.

@@

The proof of this proposition is the main technical contribution of this
paper. It also represented the main difficulty in our work. 
The remainder of this section is organized as follows: first we will
define the explicit coercion derivation algorithm which will permit to
put proof obligations in the terms, then we will define the
interpretation of terms which is mutually recursive with the precedent
algorithm. Finally we will present a proof sketch of the aforementionned
proposition.

@@

\subtis[Coercion derivation]
